---
title: "Sharpe lab - Tumour RNASeq"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 4
    fig_caption: true
    fig_width: 8
    fig_height: 6
author: "Meeta Mistry"
---

```{r setup, echo=FALSE}
 
# Setup report details
clientname="Vikram Juneja"
clientemail="vikram.juneja@gmail.com "
lablocation="Sharpe Lab- HMS"
analystname="Meeta Mistry"
analystemail="mmistry@hsph.harvard.edu"
```

RNA-Seq analysis for `r clientname` (`r clientemail`) at `r lablocation`. Contact `r analystname` (`r analystemail`) for additional details. Request from client was:     

Looking at PD1-sensitive (MC38) versus PD1-insensitive (B16): What pathways may compensate for the loss of PD-1/PD-L1 signaling in PD-1 -/- mice? What pathways do PD-1 insensitive tumor cells take advantage of?                                                 

  * Group 1: WT versus PD-1-/- (MC38) (triplicate)
  * Group 2: WT versus PD-1-/- (B16) (triplicate)
  * Group 3: MC38 in vitro (triplicate)
  * Group 4: B16 in vitro (triplicate)
  
  
```{r workflow, fig.align='center', echo=FALSE, warning=FALSE, message=FALSE}
require(png)
require(gridExtra)
img <- readPNG("additional files/workflow.png")
grid.raster(img)
```

## Workflow:   
  * run fastq files through [bcbio](https://bcbio-nextgen.readthedocs.org/en/latest/index.html) on Orchestra
  * assess QC [report link](./qc-summary/qc-summary.html)
  * simple DEG list for the WT and KO comparison within each cell line

## Setup
### Bioconductor and R libraries used

```{r libraries, echo=TRUE}
loadlibs <- function(){
library(ggplot2)
library(reshape)
library(gplots)
library(RColorBrewer)
library(CHBUtils)
library(ggdendro)
library(grid)
library(gridExtra)
library(limma)
library(DESeq2)
library(Biobase)
library(genefilter)
library(splitstackshape)
library(GOstats)
library(edgeR)
library(CHBUtils)
}
suppressPackageStartupMessages(loadlibs())
```

### Set variables
```{r directories, echo=TRUE}
baseDir=getwd()
dataDir=paste(baseDir, "/data", sep="")
resultsDir=paste(baseDir, "/results", sep="")
metaDir=paste(baseDir, "/meta", sep="")

heatcolors.1 <- rev(brewer.pal(6, "YlOrRd"))
heatcolors.2 <- colorRampPalette(brewer.pal(9, "GnBu"))(100)
cbPalette <- cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", 
                            "#D55E00", "#CC79A7", "#000000")
```
  
## RNA-Seq
### Load data

```{r loadcounts}
data <- read.delim(file.path(dataDir, 'combined.counts'), header=T, sep="\t", row.names=1, as.is=T) 
meta <- read.delim(file.path(metaDir, 'project2.csv'), header=T, sep=",", row.names=1)
meta$sampletype <- factor(paste(meta$celline, "_", meta$genotype, sep=""))
shortname = sapply(row.names(meta), function(x){
                  s <- unlist(strsplit(x, "_"))
                  newname <- paste(s[5], "_", s[6], sep="")}, USE.NAMES=F)
meta$shortname <-shortname

# Reorder data to match meta
ord.names <- sapply(rownames(meta), function(x){
                            s <- strsplit(x, ".", fixed=T)[[1]][1] 
                            grep(s, colnames(data))})

data <- data[,as.numeric(ord.names)]
colnames(data) <- row.names(meta)

# Create eset object
eset <- new("ExpressionSet", exprs=as.matrix(data))
pData(eset) <- meta
annot <- read.delim(file.path(dataDir, 'annotated_combined.counts'), sep="\t", header=T, row.names=1)
fData(eset) <- data.frame(symbol=annot$symbol, row.names=row.names(annot), stringsAsFactors = FALSE)
```
  
### Quality control
From the PCA, samples separate into four distinct group which corresponds to each cell line in vitro and exvivo. The KO and WT do not separate out, but they do tend to cluster together

```{r qcseq, echo=FALSE, fig.align='center', fig.width=22, warning=FALSE, message=FALSE}
# PCA plot of first and second PCs
dge <- DGEList(counts=data)
dge <- calcNormFactors(dge)
normalized_counts = cpm(dge, normalized.lib.sizes=TRUE)
colnames(normalized_counts) <- meta$shortname
p1 <- mds(normalized_counts, condition=pData(eset)$sampletype, k=length(colnames(data)) - 1)

# Dendrogram
  meta.x <- meta
  row.names(meta.x) <- colnames(counts)
  myDist <- dist(t(normalized_counts))
  myTree <-hclust(myDist)
  dhc <- as.dendrogram(myTree)
  ddata <- dendro_data(dhc, type="rectangle")
  ddata$labels <- merge(ddata$labels, meta.x, by.x="label", by.y="shortname")
p2 <- ggplot(segment(ddata)) +
    geom_segment(aes(x=x, y=y, xend=xend, yend=yend)) +
    theme_dendro() +
    geom_text(data=label(ddata), aes(x=x, y=y, label=label, color= label(ddata)[['sampletype']], 
                                     hjust=-0.1), size=4) +
    coord_flip() + scale_y_reverse(expand=c(0.2, 50)) +
    theme(axis.text.x=element_blank(),
          axis.text.y=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          legend.title=element_blank())

grid.arrange(p1, p2, ncol=2)
```

### DESeq2: Finding genes differentially expressed between WT and KO

Starting from count data; create a DESeq object. The design formula (~) is used to estimate the dispersions and to estimate the log2 fold changes of the model. DESeq2 will estimate the library sizes with "estimateSizeFactors". However, this can only take into account tags that have contributed to your cluster definitions (genes). Imagine that none of your clusters are differentially expressed, but due to some biological or technical artifact, the reads that do not map to clusters are more abundant in one than in the other sample. A normalization based on all counts will skew the clusters in order to account for the non-clusters. Finally, There are three levels of filtering applied in DESeq2.

1. If within a row, all samples have zero counts, the baseMean column will be zero, and the log2 fold change estimates, p-value and adjusted p-value will all be set to NA.
2. If a row contains a sample with an extreme count outlier then the p-value and adjusted p-value are set to NA. These outlier counts are detected by [Cook’s distance](http://en.wikipedia.org/wiki/Cook%27s_distance)
3. If a row is filtered by automatic independent filtering, based on low mean normalized count, then only the adjusted p-value is set to NA. 

```{r deseq2-setup, fig.width=20, warning=FALSE, message=FALSE}

# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = exprs(eset), colData = pData(eset), design = ~ sampletype)
dds <- DESeq(dds)

# Matrix of pseudocounts for downstream visualization: two methods
rld <- rlog(dds)

# Extract data of specified contrasts for B16
# resultNames(dds) # uncomment to check names for contrast setup
contrast <- list("sampletypeB16.GFP_KO", "sampletypeB16.GFP_WT")
resB16 <- results(dds, contrast=contrast)

# Extract data of specified contrast for MC38 cells
contrast <- list("sampletypeMC38.GFP_KO", "sampletypeMC38.GFP_WT")
resMC38 <- results(dds, contrast=contrast)


# Plot p-value distributions
df <- data.frame(resB16[4:39017,])
p1 <- ggplot(df, aes(x=pvalue)) + geom_histogram(binwidth=.05) +
      scale_x_continuous(limits=c(0,1)) +
      ggtitle("B16 cells") +
    theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5)),
        axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.25))) 

df <- data.frame(resMC38[4:39017,])
p2 <- ggplot(df, aes(x=pvalue)) + geom_histogram(binwidth=.05) +
      scale_x_continuous(limits=c(0,1)) +
      ggtitle("MC38 cells") +
    theme(legend.position = "none",
        plot.title = element_text(size = rel(1.5)),
        axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.25))) 

grid.arrange(p1, p2, ncol=2)

```

### Significant genes
The p-value distributions indicate that there is probabaly little to no DE with the B16 cells (based on the almost uniform distribution), but there might be something in the MC38 comparison. Significant genes were identified from each comaprison using the threshold of padj < 0.05 and an absolute FC > 2. For B16 cells only four genes passed this threshold, and for MC38 cells no significant genes were found. A less stringent alternative is to see how many genes are found using the adjusted p-value criteria on its own (p < 0.05). With this we find 8 genes differntially expressed in B16 cells and 489 genes in the MC38 cells.

There was no overlap found between the two. Gene lists are provided in the links below with Ensembl ID, Gene symbol and RPKM values in each sample. The eight genes from B16 are displayed in the table below.                            

  * Group 1: [B16 comparison](./results/resultsB16.csv)  
  * Group 2: [MC38 comparison](./results/resultsMC38.csv)

```{r siggenes, fig.align='center', echo=FALSE, eval=FALSE}
# Get significant genes from each comparison
resultsAll <- list(resB16, resMC38)
p.cutoff <- 0.05
fc <- 2
sigmat <- lapply(resultsAll, function(x){
            getGenes <- data.frame(x[4:39017,])
            getGenes <- getGenes[which(getGenes$baseMean > 0),]
            gene.FDR <- row.names(getGenes)[which(getGenes$padj < p.cutoff)]
            return(gene.FDR)})

# get gene lists
fpkm <- read.delim(file.path(dataDir, "combined.fpkm"), header=T, sep="\t", row.names=1)
out <- fpkm[sigmat[[2]],]
cols <- c("GeneSymbol", meta[colnames(out), 'shortname'])
an.out <- match(sigmat[[2]], row.names(annot))
out <- cbind(annot$symbol[an.out], out)
names(out) <- cols
write.table(out, "results/reusltsMC38.csv", sep=",", quote=F)

```

```{r B16-table, echo=FALSE, fig.align='center'}

out <- data.frame(resB16[sigmat[[1]],])
an.out <- match(sigmat[[1]], row.names(annot))
out <- cbind(annot$symbol[an.out], out)
colnames(out)[1] <- "geneSymbol"
kable(out, format='html')
```

